Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection)

### 2017-08-12 code/Rmd/ALL_REPROCESSED.Rmd
Minor changes

covariates$ROSMAP = covariates$ROSMAP %>%
  dplyr::mutate(SampleID = as.character(SampleID),
                BrainRegion = 'DLPFC',
                BrainRegion.Diagnosis = factor(paste(BrainRegion, Diagnosis, sep = '.')),
                msex = factor(msex, levels = c('1', '0'), labels = c('1'='MALE','0'='FEMALE')),
                Batch = factor(Batch)) %>%
  dplyr::rename(Gender = msex, RIN = RINcontinuous, AOD = age_death, PMI = pmi) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis, Batch, Gender, AOD, PMI, RIN, PCT_PF_READS_ALIGNED, 
                PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)
covariates$ROSMAP$individualIdentifier = factor(paste0('ROSMAPpseudoid',1:dim(covariates$ROSMAP)[1]))

covariates$MSBB = covariates$MSBB %>%
  dplyr::mutate(SampleID = as.character(SampleID),
                individualIdentifier = factor(paste0('BB',individualIdentifier)),
                SEX = factor(SEX, levels = c('M', 'F'), labels = c('M'='MALE','F'='FEMALE')),
                batch = factor(batch),
                PMI = PMI/60) %>%
  dplyr::rename(Batch = batch, Gender = SEX) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis, individualIdentifier, Batch, Gender, AOD, PMI, RIN, 
                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)

covariates$MAYO = covariates$MAYO %>%
  dplyr::mutate(SampleID = as.character(SampleID),
                Donor_ID = factor(paste0('MAYO', Donor_ID)),
                Gender = factor(Gender, levels = c('M', 'F'), labels = c('M'='MALE','F'='FEMALE')),
                Batch = factor(Source)) %>%
  dplyr::rename(individualIdentifier = Donor_ID, AOD = AgeAtDeath) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis, individualIdentifier, Batch, Gender, AOD, PMI, RIN, 
                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)

@@

```{r preprocess.data1, results='asis'}
# Filter OTHER samples 
covariates = rbindlist(covariates, use.names = T, fill = T, idcol = 'Study') %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  dplyr::filter(Diagnosis != 'OTHER') %>%
  dplyr::mutate(BrainRegion = factor(BrainRegion, levels = c('DLPFC', 'FP', 'IFG', 'PHG', 'STG', 'CER', 'TCX')),
                BrainRegion.Diagnosis = factor(paste(BrainRegion, Diagnosis, sep = '.')))

@@

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("individualIdentifier", "Study", "BrainRegion.Diagnosis", "BrainRegion", "Diagnosis",
                      "Gender", "Batch")
ContCovariates <- c("AOD", "PMI", "RIN", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES",
                    "PCT_RIBOSOMAL_BASES")

@@

### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
Differential expression is performed on the BrainRegion.Diagnosis x Gender variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=7, fig.width=24}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$BrainRegion.Diagnosis.Gender = factor(paste(COVARIATES$BrainRegion.Diagnosis,
                                                       COVARIATES$Gender,
                                                       sep = '_'))

>>>>

covariates$ROSMAP = covariates$ROSMAP %>%
  dplyr::mutate(SampleID = as.character(SampleID),
                BrainRegion = 'DLPFC',
                BrainRegion.Diagnosis = factor(paste(BrainRegion, Diagnosis, sep = '.')),
                msex = factor(msex, levels = c('1', '0'), labels = c('1'='MALE','0'='FEMALE')),
                Batch = factor(Batch),
                BrainRegion.ApoE = paste('DLPFC', apoe_genotype, sep = '.')) %>%
  dplyr::rename(Gender = msex, RIN = RINcontinuous, AOD = age_death, PMI = pmi) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis, Batch, Gender, AOD, PMI, RIN, PCT_PF_READS_ALIGNED, 
                PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES, BrainRegion.ApoE)
covariates$ROSMAP$individualIdentifier = factor(paste0('ROSMAPpseudoid',1:dim(covariates$ROSMAP)[1]))

covariates$MSBB = covariates$MSBB %>%
  dplyr::mutate(SampleID = as.character(SampleID),
                individualIdentifier = factor(paste0('BB',individualIdentifier)),
                SEX = factor(SEX, levels = c('M', 'F'), labels = c('M'='MALE','F'='FEMALE')),
                batch = factor(batch),
                PMI = PMI/60) %>%
  dplyr::rename(Batch = batch, Gender = SEX) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis, individualIdentifier, Batch, Gender, AOD, PMI, RIN, 
                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, 
                PCT_RIBOSOMAL_BASES, BrainRegion.ApoE)

covariates$MAYO = covariates$MAYO %>%
  dplyr::mutate(SampleID = as.character(SampleID),
                Donor_ID = factor(paste0('MAYO', Donor_ID)),
                Gender = factor(Gender, levels = c('M', 'F'), labels = c('M'='MALE','F'='FEMALE')),
                Batch = factor(Source)) %>%
  dplyr::rename(individualIdentifier = Donor_ID, AOD = AgeAtDeath) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis, individualIdentifier, Batch, Gender, AOD, PMI, RIN, 
                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, 
                PCT_RIBOSOMAL_BASES, BrainRegion.ApoE)
levels(covariates$MAYO$BrainRegion.Diagnosis) = c("CER.AD", "CER.CONTROL", "CER.OTHER", "TCX.AD", "TCX.CONTROL", "TCX.OTHER")

@@

```{r preprocess.data1, results='asis'}
# Filter OTHER samples 
covariates = rbindlist(covariates, use.names = T, fill = T, idcol = 'Study') %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  dplyr::filter(Diagnosis != 'OTHER') %>%
  dplyr::mutate(BrainRegion = factor(BrainRegion, levels = c('DLPFC', 'FP', 'IFG', 'PHG', 'STG', 'CER', 'TCX')),
                BrainRegion.Diagnosis = factor(paste(BrainRegion, Diagnosis, sep = '.')),
                BrainRegion.ApoE = factor(as.character(BrainRegion.ApoE)))

@@

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("individualIdentifier", "Study", "BrainRegion.Diagnosis", "BrainRegion", "Diagnosis",
                      "Gender", "Batch", "BrainRegion.ApoE")
ContCovariates <- c("AOD", "PMI", "RIN", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES",
                    "PCT_RIBOSOMAL_BASES")

@@

### Differential expression analysis (with BrainRegion.ApoE as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.ApoE', setdiff(adjust.covars, 'BrainRegion.Diagnosis')), drop = F], 
                         Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Fit contrast
cntr = c()
for (region in c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  tmp = combn(paste0('BrainRegion.ApoE', region, '_', c('0', '1', '2')),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.ApoE','',Comparison),
                Study = 'ALL',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

all.diff.exp = c(all.diff.exp, list(ApoE = DE))
all.fit = c(all.fit, list(ApoE = FIT))
```

### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
Differential expression is performed on the BrainRegion.Diagnosis x Gender variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp3, fig.height=7, fig.width=24}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$BrainRegion.Diagnosis.Gender = factor(paste(COVARIATES$BrainRegion.Diagnosis,
                                                       COVARIATES$Gender,
                                                       sep = '_'))
